---
title: "Astronomical data project"
author: "Radosław Kluczewski"
date: "26.01.2022"
output: 
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

# Introduction to Asteroids data

In this short data project using R language Firstly it has been loaded data form 
AstroStatistics page. After that, the basic statistics are presented below:

```{r include = FALSE}
# Load the required libraries
library(nortest)
library(outliers)
library(MASS)

# Set working directory. If you don't know the path, enter pwd in the terminal to find your path
 setwd("~/GitHub/R-project/main")
# Load data with the file from data folder
asteroids <- read.table("./data/asteroids.txt", header = T)

# Sort data using second column
asteroids_sorted <- asteroids[order(asteroids$Dens),]

# Assign columns
names <- asteroids_sorted[, 1]
densinity <- asteroids_sorted[, 2]
error <- asteroids_sorted[, 3]
```

```{r echo=FALSE}
summary(asteroids)
```

The standard deviation $\sigma$ has also been calculated and will be used in 
the following median standard error formula:

```{=tex}
\begin{equation}
    SE_{me}=\frac{1,253\cdot \sigma}{\sqrt{N}}.
    \label{wz1}
\end{equation}
```
The results are presented below after substituting them to the above formula, 
where for the dens and err columns:

```{r include=FALSE}
# Calculate standard deviation
standard_deviation_densinity <- sd(densinity)
standard_deviation_err <- sd(error)

# Calculate median standard error
se_densinity <- (standard_deviation_densinity * 1.253) / sqrt(26)
se_error <- (standard_deviation_err * 1.253) / sqrt(26)
```

```{r echo=FALSE}
print(se_densinity)
print(se_error)
```

The figures below show the distribution of the data where errors are also included:

```{r figure 1, echo=FALSE, warning=FALSE}
graph_data <-
  dotchart(
    densinity,
    labels = names,
    cex = 0.9,
    xlab = expression(Densinity ~ (g / cm ^ 3))
  )
```

A drawing showing the density of individual asteroids. As can be seen, the highest 
density of data occurs around densities with values from 1 to 3. Very high densities 
were observed for the remaining data, which may raise suspicions. In order to check 
if the high-density data is true, the graph with errors was drawn, which is presented below:

```{r figure 2, echo=FALSE, warning=FALSE}
graph_errors <-
  plot(
    densinity,
    ylim = c(0, 8),
    xlab = "Asteroids",
    ylab = expression(Densinity ~ (g / cm ^ 3)),
    pch = 20
  ) + grid()

number <- seq(1, length(densinity))
graph_with_all <- segments(number, densinity + error, number, densinity - error)
```

The chart above shows very high uncertainties for large points densities. 
So, these points may not necessarily be so large density values. The last graph 
drawn is the comparison distribution of asteroid density to the distribution of 
errors in their measurements. This chart allows you to check whether the data is 
of similar density or not. In addition, it allows you to determine the size of 
the errors compared to the measurements densities and whether these errors are 
similar or outlier.

```{r figure 3, echo=FALSE, warning=FALSE}
graph_comprasion <-
  boxplot(
    asteroids[, 2:3],
    varwidth = T,
    notch = T,
    xlab = "Asteroids",
    ylab = "Densinity
",
pars = list(
  boxwex = 0.3,
  boxlwd = 1.5,
  whisklwd = 1.5,
  staplelwd = 1.5,
  outlwd = 1.5,
  font = 2
)
  )
```

As we can read from the drawing, the data has a similar density, a errors are 
similar to each other.

To check whether the data can be described by a normal distribution two tests 
were decided on:

**Kolmogorov-Smirnov test:**

```{r echo=FALSE}
lillie.test(densinity)
lillie.test(error)
```

This test is one of the best-known collective tests on normality, where he tests 
the null hypothesis that the decomposition of our variable is close to normal. 
As a test statistic we accept:

```{=tex}
\begin{equation}
  D_n=\sup_x|F_0(x)-S_n(x)|,
  \label{wz2}
\end{equation}
```
where $s_n(x)$ is the empirical difference which is based on the sample order 
as follows:
```{=tex}
\begin{equation}
  F_n(x)=\frac{1}{n}\sum_{i=1} ^ n I_{X_i \leqslant x},
  \label{wz3}
\end{equation}
```
where: $X_i$ is the value of the X variable for the i-th observation. 
$I\_{X_i \leqslant x}$ is a characteristic function of the receiving set 
value one if $X_i \leqslant x$ and zero otherwise. Big the statistics values 
indicate a large discrepancy in the distribution function $F_0$ and $S_n$, 
so we construct a right-hand critical area. If the value of $D_n$ falls into the
critical area, we reject the null hypothesis on adopted significance level.

Assuming the significance level of $\alpha = 0.05$ and comparing go
with the obtained p-value, we can say whether the hypothesis can be
accept or reject. For density, the hypothesis is correct, because this
value is much greater than the assumed value $\alpha$, while for
uncertainty we reject the hypothesis because value p-value is much
smaller than $\alpha$.

**Shapiro-Wilk test:**

```{r echo=FALSE}
shapiro.test(densinity)
shapiro.test(error)
```

This test verifies that the sample is from a population with a
distribution normal. The test statistic is:
```{=tex}
\begin{equation}
  W=\frac{(\sum_{i=1} ^ n a_i x_{(i)}) ^ 2}{\sum_{i = 1} ^ n(x_i - \overline{x}) ^ 2},
  \label{wz4}
\end{equation} 
```
where: $x_i$ is $i$ - the smallest number in the sample, $\overline{x}$ 
is the sample mean, while \$ a_i \$ is expressed as pattern: 
$(a_1 ,\ldots ,a_n)=\frac{m ^ T V ^ {-1}}{C}$. 
$C$ from the previous one of the formula is the norm of the vector, which is
described by the formula: $C=||m ^ T V ^ {-1}||=\sqrt{m ^ T V ^ {-1} V ^ {-1} m}$, 
where $m=(m_1, \ldots, m_n) ^ T$ includes observations for non-decreasing values.

The null hypothesis is derived from samples from a normally distributed
population. For the previously adopted significance level, we can say
that We can assume that the hypothesis for the density of asteroids is
correct, while the hypothesis for uncertainty must be rejected because
it is much less p-value.

Based on the Dixon and Grubb tests, it was checked whether in data,
there are outliers. Outliers it points where the measurement may be the
result of a gross error. The first test is the Dixon test, the results
of which were presented below:

```{r echo=FALSE}
dixon.test(densinity)
```

The Dixon Test is a test to check whether a sample contains data
resulting from committing a gross error. The test statistic is: 
$Q =\mathrm {\frac{gap}{range}},$ where gap is the module from the
difference between the suspect measurement and the value of the closest
measurement. Range is the difference between the largest value of the
sample and the smallest value of z samples.

Another test for detecting a gross error with an error is the test
Grubbs, which consists in defining the $H_0:$ hypothesis of no
deviation in the dataset and $H_a:$ that is, is there a risk of a
deviation in dataset. The test statistic is defined as:

```{=tex}
\begin{equation}
  G=\frac{max|X_i - \overline{X}|}{\sigma},
  \label{wz5}
\end{equation}
```

where $ \overline{X} $ - mean, $ \sigma $ - deviation standard. The
Grubbs statistic is considered to be the greatest deviation from the
mean in a normally distributed set. The test results are presented as
follows:

```{r echo=FALSE}
grubbs.test(densinity)
```

Analyzing the test results, it can be concluded that we can not do any
of the points considered a gross mistake. Especially the points that
have the highest density.

# Globular clusters

```{r include=FALSE}
# Load globular clusters data from two files
galaxies_M31 <- read.table('./data/GlobClus_M31.dat', header = T)
galaxies_MW <- read.table('./data/GlobClus_MWG.dat', header = T)

# Assign variables
m_31 <- galaxies_M31[,2] 
milky_way <- galaxies_MW[,2]
```

Below are the statistics for two datasets, where the first dataset is
observations of globular clusters in the Milky Way:

```{r echo=FALSE}
summary(galaxies_MW)
```

while the second is observations of the globular clusters in M31:

```{r echo=FALSE}
summary(galaxies_M31)
```

The data distributions for which the decision was made are presented
below draw histograms for graphical representation the frequency of
occurrence of brightness in a given range.

```{r figure 4, echo=FALSE, warning=FALSE}
histogram_milky_way <- hist(milky_way)
histogram_m_31 <- hist(m_31)
```

It can be seen from the graphs that the brightness value is in both
histograms completely different. If you want to present the
distributions in one graph, you should take into account the correction
by subtracting the difference of the medians of the Milky Way and M31.
In for this purpose, the means were calculated and then subtracted from
each other:

```{r echo=FALSE}
module <- median(milky_way) - median(m_31)
print(module)
```

The result will be a similar brightness level to that observable in our galaxy. 
The results of the chart comparison remained presented below:

```{r figure 5, echo=FALSE, warning=FALSE}
graph_distribution_mw <-
  plot(
    ecdf(milky_way),
    cex.points = 0,
    verticals = T,
    main = "Empirical Distributors",
    col = "green"
  )
graph_distribution_m31 <-
  plot(
    ecdf(m_31 + module),
    cex.points = 0,
    verticals = T,
    add = T,
    main = "Empirical Distributors",
    col = "red"
  )
legend <-
  legend(
    -8,
    0.2,
    legend = c("Milky way", "M31"),
    col = c("green", "red"),
    lty = 1,
    cex = 0.8
  )

grid()
```

The result of the **module** value was compared with the result of the rank test 
Wilcoxon, which was calculated below:

```{r echo=FALSE}
wilcox.test(milky_way, m_31 + module)
```

As can be seen, the values determined by the median and the test are similar. 
Thus, it can be assumed that the median difference as a shift is satisfactorily 
sufficient..

Q-Q - a quantile-quantile plot is used, among other things, to compare or the 
data distribution follows the model. Two-dimensional graph variables represents 
the points that arise on the plot as a result two corresponding quantile values 
of their random variables decomposition. it is worth noting that if all points 
correspond with the line straight line, the random variables in the plot are described 
by the same decomposition. The procedure for creating a chart is as follows as follows:


- ordering in ascending order residues, which are empirical quantiles,
- compute quantile orders,
- compute theoretical quantiles,
- sort residuo ascending for the second dataset,
- summary of quantiles in the plot. 


The quantile graphs of the quantile are presented below:

```{r figure 6, echo=FALSE, warning=FALSE}
graph_quantil_mw_1 <-
  qqnorm(
    milky_way,
    pch = 20,
    cex.axis = 1.3,
    cex.lab = 1.3,
    main = ""
  )
graph_quantil_mw_2 <- qqline(milky_way, lty = 2, lwd = 1.5)
```

The OY axis of the above graph shows the empirical light distribution
clusters in the Milky Way, and the OX axis theoretical distribution
normal distribution. As can be seen, the empirical quantiles agree
with the theoretical model.

```{r figure 7, echo=FALSE, warning=FALSE}
graph_quantil_m31_1 <-
  qqnorm(
    m_31,
    pch = 20,
    cex.axis = 1.3,
    cex.lab = 1.3,
    main = ""
  )
graph_quantil_m31_2 <- qqline(m_31, lty = 2, lwd = 1.5)
```

Similarly as before, the OY axis of the above graph is shown
empirical distribution of the brightness of the clusters in M31, while the OX axis
theoretical distribution of the normal distribution. It can also be seen that
the theoretical model does not agree with the lower values.

Comparing the distribution of two data sets, we get the following
chart:

```{r figure 8, echo=FALSE, warning=FALSE}
graph_quantil_all <-
  qqplot(
    m_31,
    milky_way,
    pch = 20,
    cex.axis = 1.3,
    cex.lab = 1.3,
    main = "",
    xlab = "M31",
    ylab = "Milky way"
  )
```

In the above case, the axes are empirical luminance distributions
clusters in the Milky Way and in M31. In the figure above, we can
observe a non-linear trend that leads to the conclusion that the distributions
the brightness of the two galaxies is different.

In order to confirm the above conclusion, the following test should be carried out:

```{r echo=FALSE, warning=FALSE}
ks.test(milky_way, m_31 + module)
```

The test statistics have been described in the previous task. For the level
significance $\alpha = 0.05$ the hipothesis should be rejected, but because
the significance level can also be assumed at the level of $\alpha = 0.01$ no
we are sure about the correctness of rejecting the hypothesis.

# Estimators on survey data 

A consistent estimator that is unbiased and the most efficient of the mean
is the mean value. The value of the calculated mean is presented below:

```{r echo=FALSE}
height <- survey$Height 
estimator <- mean(height, na.rm = T)
print(estimator)
```

For an unknown standard deviation and an unknown average value, we use
student statistics. I am looking for the error value at 95% confidence.

```{r include=FALSE, echo=FALSE}
# Student distribution
standard_deviation_student <- sd(height, na.rm = T)
n <- length(height)
student_t <- qt(0.975, n - 1)
print(student_t)
# The confidence interval
quantil_1 <-
  qt(0.05 / 2, 237 - 1) * standard_deviation_student / sqrt(n) + estimator
quantil_2 <-
  qt(1 - 0.05 / 2, 237 - 1) * standard_deviation_student / sqrt(n) + estimator
```

Ultimately, the confidence interval is:

```{r include=FALSE, echo=FALSE}
print(quantil_1) 
print(quantil_2)
```

$$[171.1207; 173.641],$$
while the mean value together with the error from this interval is: 
```{r include=FALSE, echo=FALSE}
quantil_mean <- mean(quantil_1, quantil_2)
quantil_error <- sd(quantil_1, quantil_2)
```

$$172.38\pm 1.34 [cm].$$

In order to test the null hypothesis, i.e. whether students smoke regardless of the amount
chi-square test was used in the exercises. The statistics describing this test are:

```{=tex}
\begin{equation}
  \label{wz6}
  \chi ^ 2 = \sum_{i=1} ^ r \sum_{k=1} ^ s \frac{(n_{ik} - n \cdot p_{ik}) ^ 2}{ n \cdot p_{ik}},
\end{equation}
```

where: $n_ {ik}$ - counts corresponding to a given pair categories, r = 3 - number of smoking 
frequency categories, s = 4 - number of exercise frequency categories, 
n - sum of $n_ {ik}$ and $p_ {ik}$ which is calculated from the formula:

```{=tex}
\begin{equation}
  p_{ik} = p_{i.} p_{.k} = \frac{n_{i.}}{n} \cdot \frac{n_{.k}}{k}.
  \label{wz7}
\end{equation} 
```

In total, the statistics has 6 degrees of freedom, while the the critical area 
is on the right. The following result was obtained chi-square test:

```{r echo=FALSE, warning=FALSE}
Exer <- survey$Exer
Smoke <- survey$Smoke
Table_Smoke <- table(survey$Exer, survey$Smoke)
print(Table_Smoke)
chisq.test(Exer, Smoke)
```

It can be concluded from the test result that for the significance level 
$\alpha = 0.05$ we have no grounds to reject the null hypothesis.

# Linear Regression on faithful data

The following commands were executed to fit linear regression:

```{r echo=FALSE}
x <- faithful$eruptions
y <- faithful$waiting
regline <- lm(x ~ y)
summary(regline)

plot(x ~ y, ylab = "The length of the eruption [min]", xlab = "Waiting time for the next eruption [min]")
abline(regline, lwd = 2, col = 2)
```

Above are the statistics and the length dependence graph eruptions from waiting 
time for the next eruption along with the fitted straight using linear regression. 
Matching parameter values for straight:

```{=tex}
\begin{equation}
  y = a \cdot x + b,
  \label{wz8}
\end{equation} 
```

are, among others:

- $a = 0.0756 \pm 0.0023$,
- $b = -1.87 \pm 0.17.$

The above was used to estimate the duration of the next eruption
formula with matching coefficients, where x = 80 min. After
the substitution gave the following: $$4.18 \pm 0.36.$$

The residual plot of the regression fit is drawn below:

```{r figure 9, echo=FALSE, warning=FALSE}
plot(regline$residuals)
```

As can be seen, the applied linear regression model is correct because the data 
is scattered and does not show any trend.

**Akaike** and **Bayesian** criteria are used for purpose comparison of the models 
used with the data. Values obtained from these Criteria are estimates of the error 
and compliance of the samples with the model. Their the value can indicate which 
model is a better choice where a lower value means that the model for each data is
better. The values of these criteria for regression are presented below liner:

```{r echo=FALSE}
print(AIC(regline))
print(BIC(regline))
```

Using a 95% confidence level for the mean eruption duration for the time of 80 min, 
they were determined using the predict function predicted y values. Results of 
the action are presented below functions:

```{r echo=FALSE}
regline1 <- lm(eruptions ~ waiting, data = faithful)
time80 = predict(regline1, data.frame(waiting = c(80)), interval = "confidence")
print(time80)
```

Ultimately, the confidence interval read from the above values is: $$(4.1; 4.25).$$
